---
title: "CITE-seq data with MuData and Seurat"
author:
- name: "Danila Bredikhin"
  affiliation: "European Molecular Biology Laboratory, Heidelberg, Germany"
  email: "danila.bredikhin@embl.de"
- name: "Ilia Kats"
  affiliation: "German Cancer Research Center, Heidelberg, Germany"
  email: "i.kats@dkfz-heidelberg.de"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc_float: true
vignette: >
  %\VignetteIndexEntry{CITE-seq data with MuData and Seurat}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
collapse = TRUE,
comment = "#>"
)
```

## Introduction

CITE-seq data provide RNA and surface protein counts for the same cells. There are different workflows to analyse these data in R such as with [Seurat](https://satijalab.org/seurat/articles/weighted_nearest_neighbor_analysis.html) or with [CiteFuse](http://www.bioconductor.org/packages/release/bioc/vignettes/CiteFuse/inst/doc/CiteFuse.html). This tutorial shows how such data stored in MuData (H5MU) files can be read and integrated with Seurat-based workflows.

## Installation

The most recent `MuDataSeurat` build can be installed from GitHub:

```{r, eval = FALSE}
library(remotes)
remotes::install_github("PMBio/MuDataSeurat")
```

## Loading libraries

```{r setup, message = FALSE}
library(MuDataSeurat)
library(Seurat)
```

## Loading data

For this tutorial, we will use a subset of cells and features from the output of the [CITE-seq integration in muon](https://muon-tutorials.readthedocs.io/en/latest/cite-seq/1-CITE-seq-PBMC-5k.html).

This example dataset can be downloaded as a file in the H5MU format, which is then deserialised to create a Seurat object:

```{r}
# Download file into a temporary directory
fdir <- tempdir()
mdata_path <- file.path(fdir, "minipbcite.h5mu")
download.file(url="https://github.com/gtca/h5xx-datasets/blob/main/datasets/minipbcite.h5mu?raw=true", destfile=mdata_path, mode="wb")

pbmc <- ReadH5MU(mdata_path)
pbmc
```

Here, `pbmc` is a full-featured Seurat object that can be used in downstream analysis workflows.

## Visualising data

Importantly, data can now be plotted with Seurat.

```{r}
Idents(pbmc) <- "celltype"
DimPlot(pbmc, reduction = "WNN_UMAP")
```

### Markers

Cells can be coloured by expression level and plotted in a selected latent space:

```{r}
DefaultAssay(pbmc) <- "rna"
DimPlot(pbmc, reduction = "MOFA_UMAP", label = TRUE, repel = TRUE) + NoLegend() + 
FeaturePlot(pbmc, features = "MS4A1", reduction = "MOFA_UMAP")
```

RNA expression in different celltypes (genes in the subset were pre-selected to be celltype markers):

```{r}
rna_features <- c("IRF8", "FCGR3A", "CD14", "MS4A1", "IGHD", 
                  "KLRC2", "NKG7", "CCL5", "CD8B", "IL2RA", "IL7R")
DotPlot(pbmc, features = rna_features) + RotatedAxis()
```

Naïve/memory T cell surface protein expression:

```{r}
DefaultAssay(pbmc) <- "prot"
prot_features <- c("CD3-TotalSeqB", "CD45RA-TotalSeqB", "CD45RO-TotalSeqB")
t_cells <- c("CD4+ naïve T", "CD4+ memory T", "Treg", 
             "CD8+ naïve T", "CD8+ memory T")
VlnPlot(pbmc[,pbmc@meta.data$celltype %in% t_cells], 
        features = prot_features, ncol = 3)
```

## Downstream analysis

Since `pbmc` is a Seurat object, we can use corresponding processing and analysis functions. To provide an example, we will re-calculate cell neighbourhoods using the previously computed (on the full dataset) MOFA factors and then compute a non-linear embedding of cells.

```{r}
pbmc <- FindNeighbors(pbmc, reduction = "MOFA", dims = 1:30)

pbmc <- RunUMAP(pbmc, reduction = "MOFA", dims = 1:30, reduction.key = "rUMAP_", reduction.name = "rUMAP")

DimPlot(pbmc, reduction = "rUMAP")
```

## References

- [Muon: multimodal omics analysis framework](https://www.biorxiv.org/content/10.1101/2021.06.01.445670) preprint

- [Comprehensive Integration of Single-Cell Data](https://www.cell.com/cell/fulltext/S0092-8674(19)30559-8) paper

- [mudata](https://mudata.readthedocs.io/) (Python) documentation

- muon [documentation](https://muon.readthedocs.io/) and [tutorials](https://muon-tutorials.readthedocs.io/)

## Session Info

```{r}
sessionInfo()
```
